Deep learning for quality control of surface physiographic fields using satellite Earth observations

### 2023-05-15 docs/submission/template.tex
table updates

%% 2-column papers and preprints
\documentclass[hess, manuscript]{copernicus}

@@

\newpage
\section{Constructing VESPER}\label{sec:2}
In order to rapidly assess accuracy of new surface physiography fields and if their use in the model increase predictability a neural network regression model (further in the text referred as VESPER) that can learn the mapping between a set of features x and targets y is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis), see Table 1 for the full list of variables used. The target is the satellite land surface temperature (skin temperature from MODIS Aqua Day MYD11A1 v006 collection).VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature) and VESPER's accuracy evaluated. By varying the number/type and values of the input features x to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance. \newline

@@

\begin{table}[h]
	\begin{tabularx}{\textwidth}{lX}
		\hline
		Atmospheric and surface model fields (11 fields)
		     & \textbf{Pressure}: surface pressure (\textit{sp, Pa}), mean sea level pressure (\textit{msl, Pa}), \newline 
		\textbf{Wind}: 10 metre U wind component (\textit{10u, m/s}), 10 metre V wind component (\textit{10v, m/s}), \newline 
		\textbf{Temperature}: 2 metre temperature (\textit{2t, K}), 2 metre dewpoint temperature (\textit{2d, K}), skin temperature (\textit{skt, K}), ice temperature layer 1 (the sea-ice temperature in layer 0-7 cm; \textit{istl1, K}), ice temperature layer 2 (the sea-ice temperature in layer 7-28 cm; \textit{istl2, K}), \newline 
		\textbf{Surface albedo:} forecast albedo (\textit{fal, 0-1}), \newline 
	\textbf{Snow:} snow depth (\textit{sd, m} of water equivalent)

@@

		\\
		Additional surface physiographic fields:               & Difference for all main surface physiographic fields between V15 and V20 field sets, \newline 
		Difference between V20 static lake cover and monthly varying lake cover (12 maps in total),\newline 
		Saline lake cover (the proportion of a grid-cell covered by saline inland water bodies; units: 0-1)
		 \\
		\hline
	\end{tabularx}
\caption{Input features used for training the NN model VESPER; atmospheric model fields (time varying) were kept the same in all simulations,  surface  physiographic  fields  (static)  were  updated  when  going  from  the  original  data  based  on GlobeCover2009/GLDBv1 (V15 field set) to GSWE/GLDBv3 (V20 field set); in brackets are variables description (where needed), short name (according to the GRIB parameter database) and units.
}
\label{table:definitions}
\end{table}

@@

\begin{table}[h]

@@

\end{table}

@@

\begin{figure}
	\subfloat[\label{fiig:Modistime1}]{\includegraphics[width=0.48\textwidth]{example-image-a}} \\
	\caption{At ~31km resolution (a) V20 fractional lake cover and (b) difference between V20 and V15 lake covers. Over northern latitudes inland water increase in V20 compared to V15 is due to higher resolution input source and its better satellite image recognition methodologies; inland water reduction in V20 compared to V15 is due to anthropogenic land use changes (e.g. Aral Sea) or due to use of only permanent water (e.g. Australia) which was proven to better represent inland water distribution on annual basis.
	} 
	\label{fig:example_figure_a}

\begin{figure}
	\subfloat[\label{fiig:Modistime1}]{\includegraphics[width=0.48\textwidth]{example-image-b}} \\
	\caption{At ~18km resolution mean lake depth in meters (a) V15 and (b) V20. In general lake mean depth has decreased in V20 compared to V15 due to use of mean depth indirect estimates based on geological and climate information instead of default 25 meter value over lakes without in situ observations.} 
	\label{fig:example_figure_b}
\end{figure}

@@

\begin{figure}
	\subfloat[\label{fiig:Modistime1}]{\includegraphics[width=0.48\textwidth]{example-image-c}} \\
	\caption{A brief step-by-step explanation of the LST algorithm for MYD11A1 v006 collection.} 
	\label{fig:example_figure_c}
\end{figure}

@@

\begin{figure*}
	\subfloat[\label{fig:modis_plot_1}]{\includegraphics[width=0.48\textwidth]{num_obs_map}} \\
	\subfloat[\label{fig:modis_plot_2}]{\includegraphics[width=0.48\textwidth]{MODIS_obs_error}}
	\caption{For 2019 at ~31km resolution: (a) Mean daily number of Aqua-MODIS observations mapped to each ERA5 data point. The swath of the Aqua satellite is clearly visible, with more observations over 70-60°N and 60-70°S areas as Aqua follows a polar orbit, south to north, and with less observations over Equator, complex orography areas (such as the Himalayas, the Andes and the Rocky Mountains), and the Siberian Tundra (due to increased cloud cover). (b) Average error in the Aqua-MODIS LST measurement. The raw Aqua-MODIS data at ~1km resolution provides categorical LST errors with bins ≤ 1K, 1-2K, 2-3K and > 3K. When averaging to the coarser resolution a weighted average over the 1km grid-cells is computed, where the median bin value is used, and 5K for the > 3K bin. This information helps to understand that abundant number of observation does not automatically mean high quality of LST (e.g. Australia).} 
	\label{fig:MODIS_time_error_N}
\end{figure*}
\noindent The second step in joining ERA5 and Aqua-MODIS LST data uses a GPU-accelerated k-nearest neighbours algorithm \cite{Rapids}, where “nearness” on the sphere between two points is measured via the Haversine metric i.e. the geodesic distance H, following Eq. (2):
\begin{equation}
	H = 2 \arcsin (d)
\end{equation}
where
\begin{eqnarray}
	d = \sqrt{\sin^2\left(\frac{\delta\theta}{2}\right) + \cos \theta_1 \cos \theta_2 \sin^2\left(\frac{\delta \phi}{2}\right) }
\end{eqnarray} 
for two points with coordinate latitudes $\theta_{1,2}$, longitudes $\phi_{1,2}$ and $\delta \theta = \theta_2 - \theta_1$ and  $\delta \phi = \phi_2 - \phi_1$ .

@@

 \begin{table}[h]
 	\begin{tabularx}{\textwidth}{lXXXX}
 		\toprule
 		Model & 2016 atmospheric and surface fields & 2016 main physiographic fields & 2020 main physiographic fields & Additional surface physiographic fields \\
 		\hline
 		VESPER\_V15 & \checkmark & \checkmark & - & - \\
 		VESPER\_V15X & \checkmark & \checkmark & - & \checkmark \\
 		VESPER\_V20 & \checkmark & \checkmark & \checkmark & - \\
 		VESPER\_V20X & \checkmark & \checkmark & \checkmark & \checkmark \\

 		\bottomrule
 	\end{tabularx}
 	\caption{List of input files for different VESPER versions. c.f. Table \ref{table:definitions}}
 	\label{tab:vesper_table}
 \end{table}

@@

\noindent As focus of this study are lake related fields, and lakes occupy only 1.8\% of the Earth’s surface and are distributed very heterogeneously \citep{Choulga2014}, analysis of the results here was restricted to areas where there have been significant changes in the surface physiographic fields, i.e. a change in any of the surface field when going from V15 to V20 (and to V15X or V20X) of ≥ 10\% (≥ 0.1 for fractional fields); if lake or vegetation cover changed from 0.1 in V15 to 0.3 in V20 field set this change is classified as significant. Choice of ≥ 10 \% cut-off was adopted as it proofed to be a good trade off between having a sufficient number of grid points to inspect and the strength of the effect of changing the input field. As the cut-off \% increases less points are selected even though with more severe changes, whereas when the cut-off \% decreases more points are selected but it becomes more difficult to disentangle the change in the prediction accuracy from VESPER’s training noise. Alternative cut-off \% were briefly explored, but conclusions of the results remained broadly unchanged. All grid-cells selected for the analysis can be classified according to how the surface fields are updated when going from V15 to V20 (note that categories represent a systematic and consistent update across multiple related fields, and do not include any restrictions on other surface fields apart the ones mentioned): 
\begin{itemize}
	\item \textbf{Lake Updates}. The change in the lake cover \textit{cl} and lake depth \textit{dl} are significant, but the changes in ocean and glacier \textit{glm} fractions are not. This corresponds to grid-cells where lakes have been added or removed. Lake-Ground Updates is a sub-category where additional constraint that the change in the high/low vegetation fractions \textit{cvh/cvl} are not significant is in place. This then corresponds to the exchange of lakes for bare ground, or vice versa.
	\item \textbf{Vegetation Updates.} The change in the high vegetation fraction \textit{cvh} is significant, but the change in lake cover \textit{cl} is not significant. This corresponds to grid-cells where large features like forests and woodlands have been updated, exchanged for bare ground or low vegetation.
	\item \textbf{Glacier Updates}. The change in the glacier cover \textit{glm} is significant. This corresponds to any areas where the fraction of glacier ice has been updated.
\end{itemize}
All four VESPER versions were trained on 2016 data over the entire globe, and then used to predict LST for 2019. To compute the training noise (i.e. the network variance – changes in the LST predictability due to model training) every VESPER’s version was trained four times and each time with a different seed. To assess the changes of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default) a simple metric $\delta_{\rm VM}$ following:
\begin{equation}
	\delta_{\rm VM} = MAE_{\rm VESPER\_VM} -  MAE_{\rm VESPER\_V15} = MAE \left[LST_{\rm VESPER\_VM}-LST_{\rm Aqua-MODIS}\right] - MAE \left[LST_{\rm VESPER\_V15}-LST_{\rm Aqua-MODIS}\right]
\end{equation}
where $M$ represents one of the field set versions V20, V20X or V15X, and $MAE$ is computed over the whole prediction period of 2019; e.g. V20 describes the difference between MAE of differences between VESPER\_V20 LST predictions and Aqua-MODIS LST and between MAE of differences between VESPER\_V15 LST predictions and Aqua-MODIS LST, negative V20 indicates that VESPER\_V20 LST prediction is more accurate and vice versa.

@@

\begin{table}[h]
	\begin{tabular}{ccccccccc}
		\hline
		\multirow{2}{*}{Category} & \multirow{2}{*}{Number of grid cells} & 	\multicolumn{4}{c}{Training noise of VESPER\_VM, $K$} &\multicolumn{3}{c}{$\delta_{\rm VM}$} \\  
		&&V15  & V15X & V20 & V20X &delta &delta & delta  \\
		\hline 
		Lake&1631&0 & 0 & 0 & 0 &1 &1 & 1  \\
		Lake-Ground&546&0 & 0 & 0 & 0 &1 &1 & 1  \\
		Vegetation&58&0 & 0 & 0 & 0 &1 &1 & 1  \\
		Glacier&1057&0 & 0 & 0 & 0 &1 &1 & 1  \\
		\bottomrule
	\end{tabular}
	\caption{Globally averaged differences VM between mean absolute error (MAE) of [VESPER\_VM LST predictions minus Aqua-MODIS LST] and MAE of [VESPER\_V15 LST predictions minus Aqua-MODIS LST] for 2019 at ~31km resolution (where M denotes V15X, V20, V20X field sets) per grid-cell category. Negative VM values indicate increase of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default), positive VM values indicate decrease of LST predictability and suggest the presence of erroneous information in the surface physiographic fields. Training noise values are much smaller than the variance between different VESPER’s configurations, indicating that changes in LST predictability are mainly due to changes in the surface physiographic fields. \hl{TK: TBD}}
	\label{tab:categorisation}
\end{table}

>>>>

%% 2-column papers and preprints
\documentclass[hess, twostagejnl]{copernicus}

@@

\section{Constructing VESPER}\label{sec:2}
In order to rapidly assess accuracy of new surface physiography fields and if their use in the model increase predictability a neural network regression model (further in the text referred as VESPER) that can learn the mapping between a set of features x and targets y is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis), see Table 1 for the full list of variables used. The target is the satellite land surface temperature (skin temperature from MODIS Aqua Day MYD11A1 v006 collection).VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature) and VESPER's accuracy evaluated. By varying the number/type and values of the input features x to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance. \newline

@@

\begin{table*}
	\begin{tabularx}{\textwidth}{lX}
		\hline
		Atmospheric and surface model fields (11 fields)
		     & \textbf{Pressure}: surface pressure (\textit{sp, Pa}), mean sea level pressure (\textit{msl, Pa}), \newline 
		\textbf{Wind}: 10 metre U wind component (\textit{10u, m/s}), 10 metre V wind component (\textit{10v, m/s}), \newline 
		\textbf{Temperature}: 2 metre temperature (\textit{2t, K}), 2 metre dewpoint temperature (\textit{2d, K}), skin temperature (\textit{skt, K}), ice temperature layer 1 (the sea-ice temperature in layer 0-7 cm; \textit{istl1, K}), ice temperature layer 2 (the sea-ice temperature in layer 7-28 cm; \textit{istl2, K}), \newline 
		\textbf{Surface albedo:} forecast albedo (\textit{fal, 0-1}), \newline 
	\textbf{Snow:} snow depth (\textit{sd, m} of water equivalent)

@@

		\\
		Additional surface physiographic fields:               & Difference for all main surface physiographic fields between V15 and V20 field sets, \newline 
		Difference between V20 static lake cover and monthly varying lake cover (12 maps in total),\newline 
		Saline lake cover (the proportion of a grid-cell covered by saline inland water bodies; units: 0-1)
		 \\
		\hline
	\end{tabularx}
\caption{Input features used for training the NN model VESPER; atmospheric model fields (time varying) were kept the same in all simulations,  surface  physiographic  fields  (static)  were  updated  when  going  from  the  original  data  based  on GlobeCover2009/GLDBv1 (V15 field set) to GSWE/GLDBv3 (V20 field set); in brackets are variables description (where needed), short name (according to the GRIB parameter database) and units.
}
\label{table:definitions}
\end{table*}

@@

\begin{table*}

@@

\end{table*}

@@

\begin{figure}
	\subfloat[\label{fig:cl_map}]{\includegraphics[width=0.48\textwidth]{cl_map}} \\
	\subfloat[\label{fig:cl_delta_map}]{\includegraphics[width=0.48\textwidth]{cl_delta_map}}
	\caption{At ~31km resolution (a) V20 fractional lake cover and (b) difference between V20 and V15 lake covers. Over northern latitudes inland water increase in V20 compared to V15 is due to higher resolution input source and its better satellite image recognition methodologies; inland water reduction in V20 compared to V15 is due to anthropogenic land use changes (e.g. Aral Sea) or due to use of only permanent water (e.g. Australia) which was proven to better represent inland water distribution on annual basis.} 
	\label{fig:example_figure_a}
\end{figure}

\begin{figure}
	\subfloat[\label{fig:dl_map}]{\includegraphics[width=0.48\textwidth]{dl_map}} \\
	\subfloat[\label{fig:dl_delta_map}]{\includegraphics[width=0.48\textwidth]{dl_map_diff}}
	\caption{At ~18km resolution mean lake depth in meters (a) V15 and (b) V20. In general lake mean depth has decreased in V20 compared to V15 due to use of mean depth indirect estimates based on geological and climate information instead of default 25 meter value over lakes without in situ observations.} 
	\label{fig:example_figure_b}
	\end{figure}

@@

%\begin{figure}
%	\subfloat[\label{fiig:Modistime1}]{\includegraphics[width=0.48\textwidth]{example-image-c}} \\
%	\caption{A brief step-by-step explanation of the LST algorithm for MYD11A1 v006 collection.} 
%	\label{fig:example_figure_c}
%\end{figure}

@@

\begin{figure}
	\subfloat[\label{fig:modis_plot_1}]{\includegraphics[width=0.48\textwidth]{num_obs_map}} \\
	\subfloat[\label{fig:modis_plot_2}]{\includegraphics[width=0.48\textwidth]{MODIS_obs_error}}
	\caption{For 2019 at ~31km resolution: (a) Mean daily number of Aqua-MODIS observations mapped to each ERA5 data point. The swath of the Aqua satellite is clearly visible, with more observations over 70-60°N and 60-70°S areas as Aqua follows a polar orbit, south to north, and with less observations over Equator, complex orography areas (such as the Himalayas, the Andes and the Rocky Mountains), and the Siberian Tundra (due to increased cloud cover). (b) Average error in the Aqua-MODIS LST measurement. The raw Aqua-MODIS data at ~1km resolution provides categorical LST errors with bins ≤ 1K, 1-2K, 2-3K and > 3K. When averaging to the coarser resolution a weighted average over the 1km grid-cells is computed, where the median bin value is used, and 5K for the > 3K bin. This information helps to understand that abundant number of observation does not automatically mean high quality of LST (e.g. Australia).} 
	\label{fig:MODIS_time_error_N}
\end{figure}
\noindent The second step in joining ERA5 and Aqua-MODIS LST data uses a GPU-accelerated k-nearest neighbours algorithm \cite{Rapids}, where “nearness” on the sphere between two points is measured via the Haversine metric i.e. the geodesic distance H, following Eq. (2):
\begin{equation}
	H = 2 \arcsin (d)
\end{equation}
where
\begin{eqnarray}
	d = \sqrt{\sin^2\left(\frac{\delta\theta}{2}\right) + \cos \theta_1 \cos \theta_2 \sin^2\left(\frac{\delta \phi}{2}\right) }
\end{eqnarray} 
for two points with coordinate latitudes $\theta_{1,2}$, longitudes $\phi_{1,2}$ and $\delta \theta = \theta_2 - \theta_1$ and  $\delta \phi = \phi_2 - \phi_1$ .

@@

 \begin{table*}
 	\begin{tabularx}{\textwidth}{lXXXX}
 		\toprule
 		Model & 2016 atmospheric and surface fields & 2016 main physiographic fields & 2020 main physiographic fields & Additional surface physiographic fields \\
 		\hline
 		VESPER\_V15 & \checkmark & \checkmark & - & - \\
 		VESPER\_V15X & \checkmark & \checkmark & - & \checkmark \\
 		VESPER\_V20 & \checkmark & \checkmark & \checkmark & - \\
 		VESPER\_V20X & \checkmark & \checkmark & \checkmark & \checkmark \\

 		\bottomrule
 	\end{tabularx}
 	\caption{List of input files for different VESPER versions. c.f. Table \ref{table:definitions}}
 	\label{tab:vesper_table}
 \end{table*}

@@

\noindent As focus of this study are lake related fields, and lakes occupy only 1.8\% of the Earth’s surface and are distributed very heterogeneously \citep{Choulga2014}, analysis of the results here was restricted to areas where there have been significant changes in the surface physiographic fields, i.e. a change in any of the surface field when going from V15 to V20 (and to V15X or V20X) of ≥ 10\% (≥ 0.1 for fractional fields); if lake or vegetation cover changed from 0.1 in V15 to 0.3 in V20 field set this change is classified as significant. Choice of ≥ 10 \% cut-off was adopted as it proofed to be a good trade off between having a sufficient number of grid points to inspect and the strength of the effect of changing the input field. As the cut-off \% increases less points are selected even though with more severe changes, whereas when the cut-off \% decreases more points are selected but it becomes more difficult to disentangle the change in the prediction accuracy from VESPER’s training noise. Alternative cut-off \% were briefly explored, but conclusions of the results remained broadly unchanged. All grid-cells selected for the analysis can be classified according to how the surface fields are updated when going from V15 to V20 (note that categories represent a systematic and consistent update across multiple related fields, and do not include any restrictions on other surface fields apart the ones mentioned): 
\begin{itemize}
	\item \textbf{Lake Updates}. The change in the lake cover \textit{cl} and lake depth \textit{dl} are significant, but the changes in ocean and glacier \textit{glm} fractions are not. This corresponds to grid-cells where lakes have been added or removed. Lake-Ground Updates is a sub-category where additional constraint that the change in the high/low vegetation fractions \textit{cvh/cvl} are not significant is in place. This then corresponds to the exchange of lakes for bare ground, or vice versa.
	\item \textbf{Vegetation Updates.} The change in the high vegetation fraction \textit{cvh} is significant, but the change in lake cover \textit{cl} is not significant. This corresponds to grid-cells where large features like forests and woodlands have been updated, exchanged for bare ground or low vegetation.
	\item \textbf{Glacier Updates}. The change in the glacier cover \textit{glm} is significant. This corresponds to any areas where the fraction of glacier ice has been updated.
\end{itemize}
All four VESPER versions were trained on 2016 data over the entire globe, and then used to predict LST for 2019. To compute the training noise (i.e. the network variance – changes in the LST predictability due to model training) every VESPER’s version was in turn trained four times and each time with a different seed. To assess the changes of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default) a simple metric $\delta_{\rm VM}$ following:
\begin{equation} 
	\delta_{\rm VM} = \text{MAE}_{\rm VESPER\_VM} - \text{MAE}_{\rm VESPER\_15} 
\end{equation}

%\begin{align}
%\begin{eqnarray}
%	\delta_{\rm VM} = MAE_{\rm VESPER\_VM} -  MAE_{\rm VESPER\_V15} = MAE \left[LST_{\rm VESPER\_VM}-LST_{\rm Aqua-MODIS}\right] - MAE \left[LST_{\rm VESPER\_V15}-LST_{\rm Aqua-MODIS}\right]
%\end{eqnarray}
%\end{align}
where $M$ represents one of the field set versions V20, V20X or V15X, and $\text{MAE}$ is computed over the whole prediction period of 2019; e.g. V20 describes the difference between MAE of differences between VESPER\_V20 LST predictions and Aqua-MODIS LST and between MAE of differences between VESPER\_V15 LST predictions and Aqua-MODIS LST, negative V20 indicates that VESPER\_V20 LST prediction is more accurate and vice versa.

@@

\begin{table*}
	\begin{tabular}{ccccccccc}
		\hline
		\multirow{2}{*}{Category} & \multirow{2}{*}{Number of grid cells} & 	\multicolumn{4}{c}{Training noise of VESPER\_VM, $K$} &\multicolumn{3}{c}{$\delta_{\rm VM}$} \\  
		&&V15  & V15X & V20 & V20X & V15X &V20 & V20X  \\
		\hline 
		Lake&1631&0.009 & 0.008 & 0.007 & - &- &-0.37 & -  \\
		Lake-Ground&546&0.02 & 0.03 & 0.02 & - &- &-0.83 & -  \\
		Vegetation&58&0.06 & 0.10 & 0.13 & - & -&+0.03 & -  \\
		Glacier&1057& 0.07& 0.09 & 0.08 & - &-&-0.23  & -  \\
		\bottomrule
	\end{tabular}
	\caption{Globally averaged differences VM between mean absolute error (MAE) of [VESPER\_VM LST predictions minus Aqua-MODIS LST] and MAE of [VESPER\_V15 LST predictions minus Aqua-MODIS LST] for 2019 at ~31km resolution (where M denotes V15X, V20, V20X field sets) per grid-cell category. Negative VM values indicate increase of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default), positive VM values indicate decrease of LST predictability and suggest the presence of erroneous information in the surface physiographic fields. Training noise values are much smaller than the variance between different VESPER’s configurations, indicating that changes in LST predictability are mainly due to changes in the surface physiographic fields. \hl{TK: TBD}}
	\label{tab:categorisation}
\end{table*}
